---
title: "StarBASESimulator_Univariate"
author: "Philip Freda, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Description

StarBASE style QTL simulator with outer-loop Tjur's R2 calibration.

## Tidy Up

```{r, include=FALSE}
while (length(dev.list()) > 0) {dev.off()} 
rm(list = ls()) 
cat("\014")
```

## Load Required Libraries

```{r libraries}
packages <- c("dplyr", "ggplot2", "gridExtra")
invisible(lapply(packages, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))}))
```

## Set Working Directory

```{r}
base_dir <- "~/Library/CloudStorage/OneDrive-Cedars-SinaiHealthSystem/StarBASE-GP/Classification"
```

#### Parameters

```{r params}
set.seed(13)
N          <- 3000           # number of individuals
tNrSNP     <- 10000          # total SNPs
cNrSNP     <- 5              # causal SNPs
n_chr      <- 20             # number of chromosomes
snp_per_chr<- tNrSNP / n_chr # SNPs per chromosome
maf_range  <- c(0.3, 0.4)    # minor allele freq range
noise_sd   <- 0.0            # residual noise on logit scale
case_ratio <- 0.5            # fraction of cases
max_outer  <- 20             # max outer iterations for convergence
max_inner  <- 5000           # max inner iterations for hill-climb

# Inheritance models
inheritance_models <- list(
  Additive      = c(0.0, 0.5, 1.0),
  Subadditive   = c(0.0, 0.25, 1.0),
  Superadditive = c(0.0, 0.75, 1.0),
  Dominant      = c(0.0, 1.0, 1.0),
  Recessive     = c(0.0, 0.0, 1.0),
  Heterosis     = c(0.0, 1.0, 0.0),
  Underdominant = c(0.5, 0.0, 1.0),
  Overdominant  = c(0.0, 1.0, 0.5))

qtl_models   <- c("Additive", "Dominant", "Recessive", "Superadditive", "Heterosis")
tjur_targets <- c(0.01, 0.01, 0.01, 0.01, 0.01)
tjur_tol     <- 0.002 # +/- tolerance for Tjur's R2
```

#### Simulate Genotypes

```{r genotype simulation}
genos     <- matrix(NA, nrow = N, ncol = tNrSNP)
snp_names <- character(tNrSNP)
idx       <- 1
for (chr in 1:n_chr) {
  for (j in 1:snp_per_chr) {
    maf <- runif(1, maf_range[1], maf_range[2])
    raw <- rbinom(N, 2, maf)
    genos[, idx] <- ifelse(raw == 0, 0.0, ifelse(raw == 1, 0.5, 1.0))
    snp_names[idx] <- paste0(chr, ".", sample(1e5:1e7,1))
    idx <- idx + 1}}
colnames(genos) <- snp_names

# Sample causal QTLs
qtl_indices    <- sample(seq_len(tNrSNP), cNrSNP)
qtl_snp_names  <- colnames(genos)[qtl_indices]
```

#### Helper: raw Tjur's R2 for simulator beta calibration

```{r beta calibration helper}
tjur_r2 <- function(geno, beta, noise_sd, case_ratio) {
  eta  <- geno * beta + rnorm(length(geno), 0, noise_sd)
  prob <- 1 / (1 + exp(-eta))
  n_c  <- round(length(prob) * case_ratio)
  bin  <- integer(length(prob))
  bin[order(prob, decreasing = TRUE)[1:n_c]] <- 1
  mean(prob[bin == 1]) - mean(prob[bin == 0])}
```

#### Calibrate initial betas via uniroot so each QTL's raw Tjur's R2 ≈ target

```{r beta calibration}
qtl_betas <- numeric(cNrSNP)
for (i in seq_along(qtl_indices)) {
  g_raw  <- genos[, qtl_indices[i]]
  target <- tjur_targets[i]
  f_beta <- function(b) tjur_r2(g_raw, b, noise_sd, case_ratio) - target
  qtl_betas[i] <- uniroot(f_beta, c(0, 50))$root}
```

#### Precompute for hill-climb

```{r hill climb precompute}
g_list   <- lapply(qtl_indices, function(i) genos[, i])
exp_norm <- lapply(qtl_models, function(m) {
  enc <- inheritance_models[[m]]
  (enc - min(enc)) / diff(range(enc))})
exp_diff <- lapply(exp_norm, diff)
```

#### Helper: compute regression-based Tjur's R2

```{r Tjur's regression R2 helper}
compute_regression_tjurs <- function(pheno, betas) {
  sapply(seq_along(g_list), function(j) {
    g   <- g_list[[j]]
    eta <- g * betas[j]
    fit <- glm(pheno ~ eta, family = binomial)
    preds <- predict(fit, type = "response")
    mean(preds[pheno == 1]) - mean(preds[pheno == 0])})}
```

#### Inner function: build phenotype via threshold + hill climb for shape

```{r build phenotype}
build_pheno <- function(betas) {
  eta       <- Reduce(`+`, Map(function(g,b) g*b, g_list, betas))
  prob_base <- 1 / (1 + exp(-eta))
  n_c       <- round(N * case_ratio)
  pheno     <- integer(N)
  pheno[order(prob_base, decreasing = TRUE)[1:n_c]] <- 1
  
  score_fn <- function(p) {
    sum(sapply(seq_along(g_list), function(j) {
      obs  <- tapply(p, g_list[[j]], mean)
      obsn <- (obs - min(obs)) / diff(range(obs))
      sum((diff(obsn) - exp_diff[[j]])^2)}))}
  
  best_score <- score_fn(pheno)
  for (it in seq_len(max_inner)) {
    ci <- sample(which(pheno == 1), 1)
    ui <- sample(which(pheno == 0), 1)
    p2 <- pheno
    p2[ci] <- 0; p2[ui] <- 1
    sc <- score_fn(p2)
    if (sc < best_score) {
      pheno      <- p2
      best_score <- sc}}
  pheno}
```

#### Outer-loop: adjust betas until all regression-Tjur’s R2 ≥ target - tol

```{r outer loop adjustment}
for (iter in seq_len(max_outer)) {
  pheno <- build_pheno(qtl_betas)
  tjurs <- compute_regression_tjurs(pheno, qtl_betas)
  
  diffs <- tjurs - tjur_targets
  if (all(diffs >= -tjur_tol)) {
    message("All Tjurs >= target - tol on iter ", iter,
            " (min diff = ", round(min(diffs), 3), ")")
    break}
  
  low_idx <- which(diffs < -tjur_tol)
  qtl_betas[low_idx] <- qtl_betas[low_idx] *
    (tjur_targets[low_idx] / tjurs[low_idx])
  message(sprintf("Iter %d: bumped %d betas (min diff = %0.3f)",
                  iter, length(low_idx), min(diffs)))}

# push final phenotype into global scope
binary_pheno <- pheno
```

#### Final checks & summary

```{r final checks}
final_shape_sc <- sum(sapply(seq_along(g_list), function(j) {
  obs  <- tapply(binary_pheno, g_list[[j]], mean)
  obsn <- (obs - min(obs)) / diff(range(obs))
  sum((diff(obsn) - exp_diff[[j]])^2)}))

final_tjurs <- compute_regression_tjurs(binary_pheno, qtl_betas)

summary_df <- data.frame(
  SNP            = qtl_snp_names,
  Model          = qtl_models,
  Beta           = round(qtl_betas, 3),
  Observed_Tjur2 = round(final_tjurs, 4),
  Target_Tjur2   = tjur_targets)

cat("\n[CHECK] Shape match score:", round(final_shape_sc, 4), "\n")
cat("Phenotype distribution (cases / controls):\n")
print(table(binary_pheno))

cat("\nTrue causal SNPs summary:\n")
print(summary_df)

# Sanity‐check univariate regressions in one shot
reg_results <- do.call(rbind, lapply(seq_along(qtl_indices), function(i) {
  snp_name <- qtl_snp_names[i]
  model    <- qtl_models[i]
  g        <- genos[, qtl_indices[i]]
  
  # Fit logistic regression
  fit      <- glm(binary_pheno ~ g, family = binomial)
  null_fit <- glm(binary_pheno ~ 1, family = binomial)
  coefs    <- summary(fit)$coefficients[2, ]
  ll_full  <- as.numeric(logLik(fit))
  ll_null  <- as.numeric(logLik(null_fit))
  preds    <- predict(fit, type = "response")
  
  data.frame(
    SNP               = snp_name,
    InheritanceModel  = model,
    Beta              = coefs["Estimate"],
    Pval              = coefs["Pr(>|z|)"],
    PseudoR2          = 1 - (ll_full / ll_null),
    TjurR2            = mean(preds[binary_pheno == 1]) -
      mean(preds[binary_pheno == 0]),
    stringsAsFactors = FALSE)}))

cat("\nSanity-check regression stats:\n")
print(reg_results)
```

#### Plot observed vs expected proportions

```{r inheritance plotting}
plot_list <- vector("list", length(qtl_indices))
geno_levels <- c(0, 0.5, 1)

for (i in seq_along(qtl_indices)) {
  snp   <- qtl_snp_names[i]
  model <- qtl_models[i]
  geno  <- genos[, qtl_indices[i]]
  
  obs_prop <- sapply(geno_levels, function(l) mean(binary_pheno[geno == l]))
  enc      <- inheritance_models[[model]]
  exp_prop <- (enc - min(enc)) / diff(range(enc))
  
  df <- data.frame(
    Genotype      = factor(geno_levels, levels = geno_levels),
    ObsProportion = obs_prop,
    ExpProportion = exp_prop)
  
  p <- ggplot(df, aes(x = Genotype)) +
    geom_bar(aes(y = ObsProportion, fill = Genotype),
             stat = "identity", width = 0.6) +
    geom_line(aes(y = ExpProportion, group = 1),
              linetype = "dashed", linewidth = 0.8) +
    geom_point(aes(y = ExpProportion),
               shape = 21, size = 3, fill = "black", color = "white") +
    geom_text(aes(y = ObsProportion, label = sprintf("%.2f", ObsProportion)),
              vjust = -0.5, size = 3.5) +
    geom_text(aes(y = ExpProportion, label = sprintf("E=%.2f", ExpProportion)),
              vjust = 1.5, size = 3.2) +
    scale_fill_manual(values = c("green","blue","red")) +
    ylim(0, 1) +
    labs(
      title = paste("QTL:", snp, "(", model, ")"),
      x     = "Genotype",
      y     = "Proportion") +
    theme_minimal(base_size = 12) +
    theme(legend.position = "none")  # remove legends
  
  plot_list[[i]] <- p}

grid.arrange(grobs = plot_list, ncol = ceiling(sqrt(length(plot_list))))
```

#### Full GWAS and Summary Stats

```{r gwas and summary, cache=FALSE}
# Rebuild QTL summary results for inheritance model tracking
gwas_results <- data.frame(
  SNP      = colnames(genos),
  Beta     = NA_real_,
  Pval     = NA_real_,
  PseudoR2 = NA_real_,
  TjurR2   = NA_real_)

for (i in seq_len(ncol(genos))) {
  snp_data     <- genos[, i]
  fit          <- glm(binary_pheno ~ snp_data, family = binomial)
  null_fit     <- glm(binary_pheno ~ 1,         family = binomial)
  coef_sum     <- summary(fit)$coefficients
  ll_full      <- as.numeric(logLik(fit))
  ll_null      <- as.numeric(logLik(null_fit))
  preds        <- predict(fit, type = "response")
  
  gwas_results$Beta[i]     <- coef_sum[2, "Estimate"]
  gwas_results$Pval[i]     <- coef_sum[2, "Pr(>|z|)"]
  gwas_results$PseudoR2[i] <- 1 - (ll_full / ll_null)
  gwas_results$TjurR2[i]   <- mean(preds[binary_pheno==1]) -
    mean(preds[binary_pheno==0])}

# multiple‐test correction
gwas_results$P_Bonferroni <- p.adjust(gwas_results$Pval, method = "bonferroni")
gwas_results$P_FDR       <- p.adjust(gwas_results$Pval, method = "fdr")

# Top 10 by raw p-value
top_hits <- head(gwas_results[order(gwas_results$Pval), ], 10)

# Look up true inheritance models from summary_df
# (make sure summary_df exists, with summary_df$SNP & summary_df$Model)
top_hits$InheritanceModel <- summary_df$Model[
  match(top_hits$SNP, summary_df$SNP)]

cat("\nTop 10 SNPs by raw GWAS p-value:\n")
print(top_hits[, c("SNP","InheritanceModel","Beta","Pval",
                   "PseudoR2","TjurR2","P_Bonferroni","P_FDR")])

# Which of these are true QTLs?
is_true <- top_hits$SNP %in% summary_df$SNP
cat("\nAre true QTLs among top 10 hits?\n")
print(data.frame(
  SNP               = top_hits$SNP,
  IsTrueQTL         = is_true,
  InheritanceModel  = top_hits$InheritanceModel))

# True QTLs passing Bonferroni/FDR
cat("\nTrue QTLs detected at Bonferroni P<0.05:\n")
print(intersect(summary_df$SNP,
                gwas_results$SNP[gwas_results$P_Bonferroni < 0.05]))

cat("\nTrue QTLs detected at FDR<0.05:\n")
print(intersect(summary_df$SNP,
                gwas_results$SNP[gwas_results$P_FDR < 0.05]))
```

#### Final export

```{r export}
snp_split <- do.call(rbind, strsplit(colnames(genos), split = "\\."))
snp_chr <- as.integer(snp_split[, 1])
snp_bp  <- as.integer(snp_split[, 2]) 
snp_order <- order(snp_chr, snp_bp)
genos_sorted <- genos[, snp_order]
colnames(genos_sorted) <- colnames(genos)[snp_order]

genotype_df <- as.data.frame(genos_sorted)
full_data <- cbind(genotype_df, y = binary_pheno)

write.csv(full_data, 
  file = file.path(path.expand(base_dir), 
  "simulated_classification_GWAS_1.csv"),
  row.names = FALSE)
```

